Interactive Graphics for Visually Diagnosing Forest Classifiers in R

### 2022-02-01 PPforestVizupdate.Rnw
reviewer minor comments

  \item {\em Uncertainty measure for each observation:} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class. If a case is always predicted to be the true class, there is no uncertainty about an observation. Cases that are proportionately predicted to be multiple classes indicate difficulty to classify observations. They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.

  \item {\em Variable importance:} Each model uses samples of variables. So that, the accuracy of the models can be compared when the variable is included or omitted. There are several versions of statistics that use this to provide a measure of the variable importance for prediction.

@@

  @
\caption{Visualizing the PPtree model of the crab data. The tree has three nodes (top). The density plots show the data projections at each node, colored by group (middle). The dashed vertical red line indicates the split value of each node. At node 1 the blue species is separated from orange species. Nodes 2 and 3 separate the sexes, which are more confused for the blue species. Mosaic plots of the confusion table for each split (bottom). Node 1 shows the clear split of the species, with a small number of misclassifications. Node 2 where orange females are separated from orange males indicates small number of misclassifications. Node 3 where blue females are separated from blue males, shows a larger misclassification than for orange specie.\label{pptreeviz1}}
\end{figure}

@@

The coefficients of each projection are examined to define the other two importance measures. If the variables have been standardized, the magnitude of these values indicates importance. For a single PPtree the variable importance is computed by a weighted sum of the absolute values of the coefficients across nodes, then the weights take the number of classes in each node into account($G_s$)~\citep{lee2013pptree}.

The global variable importance in a PPforest then can be defined in different ways. One way is to average the variable importance from each PPtree across all the trees in the forest. Alternatively, can be computed as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree, for more details ~\cite{da2021projection}.

Figure \ref{impotree} shows the absolute projection coefficient of the top three nodes for all the trees in a forest model. This information is displayed by a side-by-side jittered dot plot. The red dots correspond to the absolute coefficient values for the tree model of Figure \ref{pptreeviz1}. The forest was built using random samples of two variables for each node, hence there are two coefficients for each node. At node 1, BD has a high value and CW contributes much less. The scatterplot at right shows these two variables and the resulting boundary between groups that this would produce. Node 2 uses CL and RW, and RW contributes the most to the separation. The plot at right shows the boundary that is induced. Node 3 uses FL and RW, and this is a much more even contribution by the two variables.
For each tree in the forest, different decision rules are defined; the resulting boundaries on the previous plots are based on Rule 1 $= \frac{m_1}{2} + \frac{m_2}{2} $, where $m_1$ and $m_2$ are the mean of the left and right groups at each node.

@@

  @
\caption{Examining similarity between cases, using pairwise plots of multidimensional scaling into three dimensional. It can be seen that most cases are grouped closely with their class, and particularly that the two species are distinct. There is more confusion of cases between the sexes. \label{prox1}}
\end{figure}

@@

Figure \ref{tetra} shows the tetrahedron structure for the crab vote matrix shown in three pairwise views. With well-separated classes, the colored points will each concentrate into one of the vertices. This is close but not perfect, indicating some crabs are commonly incorrectly predicted.

@@

As \cite{sievert2020interactive} describes, one of the biggest difficulties for the app in order manage linking between plots is the data structure management for each widget. Each widget has it own data structure and interaction. Putting them into the structure of a shiny app facilitates access to the widget data and coordinates selections across multiple plots.

@@

This tab is designed to examine uncertainty in the classification of observations and to explore the similarity between pairs of observations. The data feeding the display is an $n\times p$ data frame, containing the original data, the model statistics generated from the full $n\times G$ vote matrix along with its generalized ternary coordinates, and the first two MDS projections of the proximity matrix.
Figure \ref{tab1} shows the arrangement of plots.
The plots in the tab are (1) a parallel coordinate plot (PCP) of the data, (2) the MDS display of the proximity matrix, (3) side-by-side jittered dotplot, and (4) generalized ternary plot of the vote matrix. Each of these plots are interactive in the sense that each one presents individual interactions (mouse-over) and they are linked so that selections in one display are propagated to other plots (clicking and selecting).

@@

The diagram in Figure \ref{tab1diag} illustrates the data pipeline \citep{BAHM88,wickham2009plumbing} for the interactive graphics in the case level tab. Solid lines indicate notifications from the source data to the plots, and dashed lines indicate notification of user action on the plot, which notifies the data source of actions to take. The data table is a reactive object that has a listener associated with it. Each of the plots is reactive and has numerous listeners. When users make selections on a plot, either by clicking or group selection, a change to the data is made in terms of an update on the selected cases. This invokes a note to other plots to re-draw themselves. The linking between plots is effectively one-to-one, based on the row id of the data. The side-by-side jittered dotplot has $n\times G$ points, but selection can only be done within a dotplot. Selecting in one of the dotplots notifies the data table of the selection which triggers a re-draw of the other dotplots. Mouseovers on the plot pull additional information about the point or line under the cursor but do not link between plots.

Two alternatives can be selected in shiny to draw the parallel coordinate plot: parallel or enhanced. Parallel draws the classic PCP and enhanced draws a modified version where variables are repeated \citep{hurley2011eulerian}.  Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variables are neighboring.

@@

This second tab in the app focuses on teasing apart the forest to examine the qualities of each tree. For each tree, information on the variable importance, the projections used at each node, and the oob error is available.  The data feeding into this tab is a list of models along with the original data frame.
The tree id is displayed when we mouse over the jittered side-by-side plot. This information is useful because, based on the accuracy, some trees could be pruned from the forest outside of the app.

@@

\subsection{Performance Comparison}
The third tab (Figure \ref{tab3}) examines the PPF fit and compares the result with a RF fit. There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) an receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) oob error by the number of trees to assess complexity, (4) overall variable importance. There is very little interaction on this tab. Users can select to focus on a subset of classes or choose the importance measure to show. Being able to focus on class can help to better understand how well the model performs across classes and the focus can be especially useful for unbalanced data. Examining the oob error by trees enables an assessment of how few trees might be used to provide an equally accurate prediction of future data.

@@

Having better tools to open up black box models will provide for better understanding the data, the model strengths and weaknesses, and how the model will performs on future data. This visualisation app provides a selection of interactive plots to diagnose PPF models. This shell could be used to make an app for other ensemble classifiers. The philosophy underlying the collection of displays is ``show the model in the data space''  explained in \citet{wickham2015visualizing}. It is not easy to do this, and to completely take this on would require plotting the model in the $p$-dimensional data space. In the simplest approach, as taken here, it means to link the model diagnostics to displays of the data. Then it is possible to probe and query to obtain a better understanding such as finding regions in the data that prove difficult to fit, and detract from the predictive accuracy, or that don't adhere to model assumptions.

@@

There are many possible extensions to the app, that could help it to be a tool for model refinement: (1) Using the diagnostics to weed out under-performing models in the ensemble; (2) Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data; (3) Problematic cases could be removed, and ensembles re-fit; (4) Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem. Working within the R environment makes all of these desires available using command line outside the app, given the unique ids of models and cases can be exported from the app.

>>>>

  \item {\em Uncertainty measure for each observation:} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class. If a case is always predicted to be the true class, there is no uncertainty about an observation. Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations. They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.

  \item {\em Variable importance:} Each model uses samples of variables. Thus, the accuracy of the models can be compared when the variable is included or omitted. There are several versions of statistics that use this to provide a measure of the variable importance for prediction.

@@

  @
\caption{Visualizing the PPtree model of the crab data. The tree has three nodes (top). The density plots show the data projections at each node, colored by group (middle). The dashed vertical red line indicates the split value of each node. At node 1 the blue species is separated from orange species. Nodes 2 and 3 separate the sexes, which are more confused for the blue species. Mosaic plots of the confusion table for each split (bottom). Node 1 shows the clear split of the species, with a small number of misclassifications. Node 2 where orange females are separated from orange males indicates small number of misclassifications. Node 3 where blue females are separated from blue males, shows a larger misclassification than for orange species.\label{pptreeviz1}}
\end{figure}

@@

The coefficients of each projection are examined to define the other two importance measures. If the variables have been standardized, the magnitude of these values indicates importance. For a single PPtree the variable importance is computed by a weighted sum of the absolute values of the coefficients across nodes, then the weights take the number of classes in each node into account~\citep{lee2013pptree}.

The global variable importance in a PPforest then can be defined in different ways. One way is to average the variable importance from each PPtree across all the trees in the forest. Alternatively, it can be computed as a weighted mean of the absolute value of the projection coefficients across all nodes in every tree, for more details ~\cite{da2021projection}.

Figure \ref{impotree} shows the absolute projection coefficient of the top three nodes for all the trees in a forest model. This information is displayed by a side-by-side jittered dot plot. The red dots correspond to the absolute coefficient values for the tree model of Figure \ref{pptreeviz1}. The forest was built using random samples of two variables for each node, hence there are two coefficients for each node. At node 1, BD has a high value and CW contributes much less. The scatterplot at right shows these two variables and the resulting boundary between groups that this would produce. Node 2 uses CL and RW, and RW contributes the most to the separation. The plot in the right shows the boundary that is induced. Node 3 uses FL and RW, and this is a much more even contribution by the two variables.
For each tree in the forest, different decision rules are defined; the resulting boundaries on the previous plots are based on Rule 1 $= \frac{m_1}{2} + \frac{m_2}{2} $, where $m_1$ and $m_2$ are the mean of the left and right groups at each node.

@@

  @
\caption{Examining similarity between cases, using pairwise plots of multidimensional scaling into a three-dimensional space. It can be seen that most cases are grouped closely with their class, and particularly that the two species are distinct. There is more confusion of cases between the sexes. \label{prox1}}
\end{figure}

@@

Figure \ref{tetra} shows the tetrahedron structure for the crab vote matrix shown in three pairwise views. With well-separated classes, the colored points will each concentrate into one of the vertices. This is close but not perfect, indicating some crabs are commonly incorrectly classified.

@@

As \cite{sievert2020interactive} describes, one of the biggest difficulties when it comes to the app managing linking between plots is the data structure management for each widget. Each widget has it own data structure and interaction. Putting them into the structure of a shiny app facilitates access to the widget data and coordinates selections across multiple plots.

@@

This tab is designed to examine uncertainty in the classification of observations and to explore the similarity between pairs of observations. The data feeding the display is an $n\times p$ data frame; containing the original data; the model statistics generated from the full $n\times G$ vote matrix along with its generalized ternary coordinates; and the first two MDS projections of the proximity matrix.
Figure \ref{tab1} shows the arrangement of plots.
The plots in the tab are (1) a parallel coordinate plot (PCP) of the data, (2) the MDS display of the proximity matrix, (3) side-by-side jittered dotplot, and (4) generalized ternary plot of the vote matrix. Each of these plots are interactive in the sense that each one presents individual interactions (mouse-over) and they are linked so that selections in one display are propagated to other plots (clicking and selecting).

@@

The diagram in Figure \ref{tab1diag} illustrates the data pipeline \citep{BAHM88,wickham2009plumbing} for the interactive graphics in the case level tab. Solid lines indicate notifications from the source data to the plots, and dashed lines indicate notification of user action on the plot, which notifies the data source of actions to take. The data table is a reactive object that has a listener associated with it. Each of the plots is reactive and has numerous listeners. When users make selections on a plot, either by clicking or group selection, a change to the data is made in terms of an update on the selected cases. This invokes a note to other plots to re-draw themselves. The linking between plots is effectively one-to-one, based on the row ID of the data. The side-by-side jittered dotplot has $n\times G$ points, but selection can only be done within a dotplot. Selecting in one of the dotplots notifies the data table of the selection which triggers a re-draw of the other dotplots. Mouseovers on the plot pull additional information about the point or line under the cursor but do not link between plots.

Two alternatives can be selected in shiny to draw the parallel coordinate plot: parallel or enhanced. Parallel draws the classic PCP and enhanced draws a modified version where variables are repeated \citep{hurley2011eulerian}.  Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variable pairs are shown next to each other once.

@@

This second tab in the app focuses on teasing apart the forest to examine the qualities of each tree. For each tree, information on the variable importance, the projections used at each node, and the oob error is available.  The data feeding into this tab is a list of models along with the original data frame.
The tree ID is displayed when we mouse over the jittered side-by-side plot. This information is useful because, based on the accuracy, some trees could be removed from the forest outside of the app.

@@

\subsection{Performance Comparison}
The third tab (Figure \ref{tab3}) examines the PPF fit and compares the result with a RF fit. There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) a receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) oob error by the number of trees to assess complexity, (4) overall variable importance. There is very little interaction on this tab. Users can select to focus on a subset of classes or choose the importance measure to show. Being able to focus on class can help to better understand how well the model performs across classes and the focus can be especially useful for unbalanced data. Examining the oob error by trees enables an assessment of how few trees might be used to provide an equally accurate prediction of future data.

@@

Having better tools to open up black box models will provide for better understanding the data, the model strengths and weaknesses, and how the model will perform on future data. This visualisation app provides a selection of interactive plots to diagnose PPF models. This shell could be used to make an app for other ensemble classifiers. The philosophy underlying the collection of displays is ``show the model in the data space''  explained in \citet{wickham2015visualizing}. It is not easy to do this, and to completely take this on would require plotting the model in the $p$-dimensional data space. In the simplest approach, as taken here, it means to link the model diagnostics to displays of the data. Then it is possible to probe and query to obtain a better understanding such as finding regions in the data that prove difficult to fit, and detract from the predictive accuracy, or that don't adhere to model assumptions.

@@

There are many possible extensions to the app, that could help it to be a tool for model refinement: (1) Using the diagnostics to weed out under-performing models in the ensemble; (2) Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data; (3) Problematic cases could be removed, and ensembles re-fit; (4) Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem. Working within the R environment makes all of these desires available using command line outside the app, given the unique IDs of models and cases can be exported from the app.
